---
title: "Data preparation and formatting."
description: |
  Data analysis for paper
author:
  - name: Paul Melloy 
    url: https://paulmelloy.github.io/
    affiliation: The University of Queensland
date: "`r Sys.Date()`"
output: distill::distill_article
---

# The influence of temperature and vapour pressure deficit on conidia germination and germ tube production in an Australian *Podosphaera xanthii* isolate.

## Compendium sections  

   1. [Title page](P_xanthii_titlepage.html)  
   2. [Data preparation and formatting](data_cleaning.html)  
   3. [The influence of temperatures on P. xanthii conidia germination](Pxanthii_temperature.html)  
   3. [The influence of VPD P. xanthii conidia germination and germtube formation](Pxanthii_vpd.html)  
   
   
The data analysed in this repository was provided by Dr Zaiton Sapak.
The final formatted data can be found in the `\cache` folder of this compendium repository.
<!-- Data pulled in and cleaned was obtained from her Google Drive folders "Chapter 5- Conidial germination and Infection" and "SAS analysis". -->  

To clean the data we will need some R libraries for handling data and data manipulation.  

```{r libraries,message = FALSE}
library(readxl) # reading microsoft excel files
library(dplyr) # Data manipulation and formatting
library(data.table) # Data manipulation and formatting
# install unpublished development version of R package `epiphytoolR`
if("epiphytoolR" %in% names(installed.packages()[, "Package"]) == FALSE) {
  remotes::install_github(repo = "https://github.com/PaulMelloy/epiphytoolR",
                          ref = "dev",
                          upgrade = "never"
                          )
}
```

<br>  

## Temperature experimental data  

### Temperature effect on germination experiment  

We will begin by reading in data from experiments that assessed the effect of temperature on conidia germination.  


```{r import_data, message = FALSE}
germ_tm <- read_xlsx("data/Chapter 5 (germination vs temperature vs time incubation at low VPD).xlsx",
                      range = "A1:N32")
```

Next we will reformat the data to make it more machine readable then save it to the `/cache` folder, which contains the formatted machine readable data.  

```{r clean_germ_tm}
# Remove unnecessary columns and rows
germ_tm <- data.table(germ_tm)[-c(1,2,7:10,15:18,23:27),-c(4,6,7,9,11,13)]

# rename leaf number column
setnames(germ_tm,
         old = c("Temperature" ,"...2"),
         new = c("i_period","leaf"))

# fill incubation period
germ_tm[13:16,i_period := "48h"]

# remove "h" and make column numeric
germ_tm[,i_period := as.numeric(gsub("h","",germ_tm$i_period))]

# fill in na's with previous non NA value
setnafill(germ_tm,
          type = "locf",
          cols = "i_period")

# reshape data to long format
germ_tm <- 
   melt(germ_tm,
     id.vars = c("i_period","leaf"),
     measure.vars = c("17oC", "19oC", "22oC", "25oC", "28oC", "31oC"),
     variable.name = "Tm",
     value.name = "germ_conidia")

# remove "oC" from temperature and make it numeric
germ_tm[, c("Tm", "leaf", "germ_conidia","non_germ_conidia") := list(as.numeric(gsub("oC","",Tm)),
                                                                     as.numeric(gsub("L","",leaf)),
                                            as.numeric(germ_conidia),
                                            100 - as.numeric(germ_conidia))]

# Change column order
setcolorder(germ_tm,
            neworder = c("Tm", "i_period", "leaf", "germ_conidia","non_germ_conidia"))
```

While not recorded in the raw data, no germination was observed at temperatures
8C or 35C. 
This is important to include for the data analysis.  

```{r}
germ_tm <- rbindlist(
   list(
      germ_tm,
      list(Tm = c(rep(8, 4 * 4),
                  rep(35, 4 * 4)),
      i_period = rep(c(12, 24, 36, 48), 
                     each = 4,
                     times = 2),
      leaf = rep(1:2, 
                 each = 2, 
                 times = 2 * 2 * 2),
      germ_conidia = rep(0, 4 * 4 * 2),
      non_germ_conidia = rep(100, 4 * 4 *2))
   ), use.names = TRUE
)
```

A quick check to look at column headers and the first 5 rows of data.  

```{r}
# inspect data
head(germ_tm)
```



```{r write_germ_tm}
# save data into cache
fwrite(x = germ_tm[order(Tm,i_period, leaf),],
       file = "cache/germination_temperature.csv")

```

<br>  

***  

### Effect of temperature on branching  

We will do the same for data collected on the influence of temperature on the number of conidia which advanced past germination to branch and possibly develop infection structures.  

```{r tm_branch}
# read in data from sas file
br_tm <- readLines("data/tempvs.branching(results).sas")[3:40]
# cut out non-data lines
br_tm <- br_tm[-2]

#write it to csv so it can be read back in cand formatted to a data.frame instead of vector.
write(br_tm, "cache/Tm_germtubes.csv")

# read in data using fread
br_tm <- fread("cache/Tm_germtubes.csv", 
              col.names = c("obs", "Tm", "germtubes","rep", "conidia"))

# Add relative humidity and temperature data
br_tm[ , c("Tm","germtubes", "obs") := list(as.numeric(gsub("oC","",Tm)),
                                         as.numeric(gsub("b","",germtubes)),
                                         NULL)]

# recalculate vpd and convert germtubes to numeric
zero_branch <- br_tm[, (100 - sum(conidia)), 
                      by = .(Tm, rep)][,germtubes :=0][, .(Tm, germtubes, rep, V1)]

setnames(zero_branch,old = "V1", new = "conidia")

# Bind in the data containing non-germinated conidia as conidia with zero germtubes
br_tm <- rbind(br_tm,zero_branch)[, conidia := round(conidia)]
```

While not recorded in the raw data, no germination was observed at temperatures
8C or 35C. 
This is important to include for the data analysis.  

```{r}
br_tm <- rbindlist(
   list(
      br_tm,
      list(Tm = c(rep(8, 4 * 2),
                  rep(35, 4 * 2)),
           germtubes = rep(c(0,1,2,3), 
                         each = 2,
                         times = 2),
           rep = rep(1:2, 8),
           conidia = rep(c(100,0,0,0), 
                         each = 2,
                         times = 2)
      )
   ), use.names = TRUE
)
```

A quick check to look at column headers and the first 5 rows of data.  

```{r}
# inspect data
head(br_tm)
```

```{r save_br_tm}
fwrite(br_tm[order(Tm,germtubes, rep),], "cache/branching_temperature.csv")
```

<br>  

***  

## Experimental data on VPD  

### Effect of VPD on germination  


```{r clean_germ_vpd}
# Import
germ_vpd <- bind_rows(
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "B63:F158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  ),
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "I63:M158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  ),
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "P63:T158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  )
)

# remove characters from temperature and time, make RH as a percentage
setDT(germ_vpd)[, c("RH", "Tm", "i_period","non_germ_n") :=
                  list(RH * 100,
                       as.numeric(gsub("C", "", Tm)),
                       as.numeric(gsub("h", "", i_period)),
                       100 - germ_n)]

# add vpd column
germ_vpd[, vpd := round(epiphytoolR::calc_vpd(RH = RH,
                                              Tm = Tm), 
                        3)]

# Change column order
setcolorder(germ_vpd,
            neworder = c("RH", "Tm", "vpd", "i_period", "rep", "germ_n", "non_germ_n"))
```


A quick check to look at column headers and the first 5 rows of data.  

```{r}
# inspect data
head(germ_vpd)
```


```{r save_g_vpd}
# write out clean data
fwrite(germ_vpd,
       file = "cache/germination_vpd.csv")
```

<br> 

***  

### Vapour pressure deficit effect on branching  
Read in the raw data.  

```{r}
# File not found 
br_vpd <-
  fread("data/220308_branching_vpd_temperature.csv",nrows = 29,header = TRUE)
```



```{r clean_br_vpd}
# Remove unnecessary columns and rows
br_vpd <- br_vpd[-c(5:10,15:24),-c(1,2,9:12,19:22,29)][-c(13)]

# rename leaf number column
setnames(br_vpd,
         old = 1:18,
         new = c("99_b1","95_b1","85_b1","75_b1","55_b1","32_b1",
                 "99_b2","95_b2","85_b2","75_b2","55_b2","32_b2",
                 "99_b3","95_b3","85_b3","75_b3","55_b3","32_b3"))

# Add temperature treatment
br_vpd[ , Tm := rep(c(22,25,28), each = 4)]

# reshape data to long format
br_vpd <-
   melt(br_vpd,
     id.vars = c("Tm"),
     value.name = "conidia",
     variable.factor = FALSE)

# split variable column into two columns vpd and germtubes and remove variable column
br_vpd[, c("RH","germtubes") := 
          tstrsplit(variable, "_b")][, c("vpd","variable") :=
                                        list(round(epiphytoolR::calc_vpd(RH = as.numeric(RH),
                                                                         Tm = as.numeric(Tm)),
                                                   3),
                                             NULL)]

# Change column order
setcolorder(br_vpd,
            neworder = c("vpd","RH", "Tm", "germtubes","conidia"))

# set correct classes and add non-germinated
br_vpd[, c("RH",
           "germtubes",
           "conidia",
           "rep") :=
              list(as.numeric(RH),
                   as.numeric(germtubes),
                   as.numeric(conidia),
                   rep(1:4,.N/4))]

# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd[, (100 - sum(conidia)), 
                      by = .(vpd,RH,Tm, rep)][,germtubes :=0][, .(vpd,RH, Tm, germtubes, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Bind and roder data containing non-germinated conidia as conidia with zero germtubes
br_vpd <- rbind(br_vpd,zero_branch)[order(Tm,germtubes, rep)]
```

Before saving the data a quick data check is made to look at column headers and 
the first 5 rows of data.  

```{r}
# inspect data
head(br_vpd)
```

```{r}
# save data into cache
fwrite(x = br_vpd,
       file = "cache/branching_VPD.csv")

```

### VPD branching dataset two  

```{r clean_br_vpd2}
br_vpd2 <- 
  read_xls("data/Chapter 5 (germination vs VPD vs temp vs RH).xls",
           sheet = "Sheet2",
           range = "AF3:AI182",
           col_names = c("vpd", "germtubes", "rep", "conidia"))

# make data.table format and add temperature and humidity
br_vpd2 <- data.table(br_vpd2)[, c("Tm","RH") := list(rep(c(22,25,28), each = 12, times = 5),
                                                            rep(c(95,85,75,55,32), each = 12*3))]
# remove "b" from germtubes and make it numeric
br_vpd2[, germtubes := as.numeric(gsub("b","",germtubes))]

# use saturated branching data from temperature experiment
br_vpd99 <- copy(br_tm)
br_vpd99[, RH := 99][, c("vpd",
                         "germtubes",
                         "rep",
                         "conidia",
                         "Tm",
                         "RH") := list(as.numeric(round(epiphytoolR::calc_vpd(RH = RH,
                                                                 Tm = Tm),
                                           3)),
                                        germtubes,
                                        rep,
                                        conidia,
                                        Tm,
                                        RH)]

br_vpd2 <- rbind(br_vpd99[
   Tm %in% unique(br_vpd2$Tm) &
      germtubes != 0,],
                 br_vpd2)

# recalculate vpd
br_vpd2[, vpd := round(epiphytoolR::calc_vpd(RH = RH,
                                                Tm = Tm),
                          3)]



# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd2[, (100 - sum(conidia)), 
                      by = .(Tm, RH,vpd, rep)][,germtubes :=0][, .(vpd,RH,Tm, germtubes, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Change column order
setcolorder(br_vpd2,
            neworder = c("vpd", "RH","Tm", "germtubes","rep","conidia"))

# Bind and roder data containing non-germinated conidia as conidia with zero germtubes
br_vpd2 <- rbind(br_vpd2,zero_branch)[order(Tm,germtubes, rep)]

```

Before saving the data a quick data check is made to look at column headers and 
the first 5 rows of data.  

```{r}
# inspect data
head(br_vpd2)
```

```{r}

# save data into cache
fwrite(x = br_vpd2,
       file = "cache/branching_VPD2.csv")

```


### More data on germtube branching under different vpd  

```{r clean_br_vpd3}
# read in data from sas file
dat <- readLines("data/vpd germtube.sas")[3:236]
# cut out non-data lines
dat <- dat[-c(2,55:58,111:114,167:170,223:226)]
#write it to csv
write(dat, "cache/vpd_germtubes.csv")

# read in data using fread
br_vpd3 <- fread("cache/vpd_germtubes.csv")

# Add relative humidity and temperature data
br_vpd3[ , c("RH","Tm") := list(rep(c(99,95,85,75,55,32),each = 36),
                                    rep(c(22,25,28),each = 12, times = 6))]

# recalculate vpd and convert germtubes to numeric
br_vpd3[, c("vpd",
                "bran") :=
              list(round(epiphytoolR::calc_vpd(RH = RH,
                                               Tm = Tm),
                         3),
                   as.numeric(gsub("b", "", bran)))]

# rearrange columns
setcolorder(br_vpd3,
            neworder = c("Obs", "RH", "Tm", "vpd", "bran", "rep", "numb")
            )

setnames(br_vpd3,
         old = c("bran", "numb"),
         new = c("germtubes","conidia"))


# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd3[, (100 - sum(conidia)), 
                      by = .(Tm, RH,vpd, rep)][,germtubes :=0][, .(vpd, RH, Tm, germtubes, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Change column order
setcolorder(br_vpd3[, c("Obs") := NULL],
            neworder = c("vpd", "RH", "Tm", "germtubes","rep","conidia"))

# Bind and roder data containing non-germinated conidia as conidia with zero germtubes
br_vpd3 <- rbind(br_vpd3, zero_branch)[order(Tm,germtubes, rep)]


```

Before saving the data a quick data check is made to look at column headers and 
the first 5 rows of data.  

```{r}
# inspect data
head(br_vpd3)
```

```{r}

fwrite(br_vpd3,
       file = "cache/branching_VPD3.csv")

```


To describe how the methods were undertaken we need to determine how many leaf discs
were inoculated and in how many runs. 
Also how many were destructively sampled in each experiment.  

```{r}
n_dat <- 
   data.table(
   germ_tm = germ_tm[ , length(unique(leaf)), by = .(Tm, i_period)][,sum(V1)],
   germ_vpd = germ_vpd[ , length(unique(rep)), by = .(vpd, i_period)][,sum(V1)],
   br_tm = br_tm[, length(unique(rep)), by = .(Tm, rep)][,sum(V1)],
   br_vpd = br_vpd[, length(unique(rep)), by = .(vpd, rep)][,sum(V1)],
   br_vpd2 = br_vpd2[, length(unique(rep)), by = .(vpd, rep)][,sum(V1)],
   br_vpd3 = br_vpd3[, length(unique(rep)), by = .(vpd, rep)][,sum(V1)]
)


t(n_dat)
```








