---
title: "The influence of temperature and vapour pressure deficit on conidia germination and germ tubes production in an Australian Podosphaera xanthii isolate."
description: |
  Data analysis for paper
author:
  - name: Paul Melloy 
    url: https://twitter.com/PaulMelloy
    affiliation: The University of Queensland
date: "`r Sys.Date()`"
output: distill::distill_article
---

The data analysed in this repository was provided by Dr Zaiton Sapak.
Data pulled in and cleaned was obtained from her Google Drive folders "Chapter 5- Conidial germination and Infection" and "SAS analysis".  

To clean the data we will need some R libraries for handling data and data manipulation.  

```{r libraries,message = FALSE}
library(readxl)
library(dplyr)
library(data.table)
library(units)
if("epiphytoolR" %in% names(installed.packages()[, "Package"]) == FALSE) {
  remotes::install_github(repo = "https://github.com/PaulMelloy/epiphytoolR",
                          ref = "dev",
                          upgrade = "never"
                          )
}
```

<br>  

## Temperature experimental data  

### Temperature effect on germination experiment  

We will begin by reading in data from experiments that assessed the effect of temperature on conidia germination.  


```{r import_data, message = FALSE}
germ_tm <- read_xlsx("data/Chapter 5 (germination vs temperature vs time incubation at low VPD).xlsx",
                      range = "A1:N32")
```

Next we will reformat the data to make it more machine readable then save it to the `/cache` folder, which contains the formatted machine readable data.  

```{r clean_germ_tm}
# Remove unnecessary columns and rows
germ_tm <- data.table(germ_tm)[-c(1,2,7:10,15:18,23:27),-c(4,6,7,9,11,13)]

# rename leaf number column
setnames(germ_tm,
         old = c("Temperature" ,"...2"),
         new = c("i_period","leaf"))

# fill incubation period
germ_tm[13:16,i_period := "48h"]
# remove "h" and make column numeric
germ_tm[,i_period := as.numeric(gsub("h","",germ_tm$i_period))]

# fill in na's with previous non NA value
setnafill(germ_tm,
          type = "locf",
          cols = "i_period")

# reshape data to long format
germ_tm <- 
   melt(germ_tm,
     id.vars = c("i_period","leaf"),
     measure.vars = c("17oC", "19oC", "22oC", "25oC", "28oC", "31oC"),
     variable.name = "Tm",
     value.name = "germ_conidia")

# remove "oC" from temperature and make it numeric
germ_tm[, c("Tm", "leaf", "germ_conidia","non_germ_conidia") := list(as.numeric(gsub("oC","",Tm)),
                                                                     as.numeric(gsub("L","",leaf)),
                                            as.numeric(germ_conidia),
                                            100 - as.numeric(germ_conidia))]

# Change column order
setcolorder(germ_tm,
            neworder = c("Tm", "i_period", "leaf", "germ_conidia","non_germ_conidia"))
```

While not recorded in the raw data, no germination was observed at temperatures
8C or 35C. 
This is important to include for the data analysis.  

```{r}
germ_tm <- rbindlist(
   list(
      germ_tm,
      list(Tm = c(rep(8, 4 * 4),
                  rep(35, 4 * 4)),
      i_period = rep(c(12, 24, 26, 48), 
                     each = 4,
                     times = 2),
      leaf = rep(1:2, 
                 each = 2, 
                 times = 2 * 2 * 2),
      germ_conidia = rep(0, 4 * 4 * 2),
      non_germ_conidia = rep(100, 4 * 4 *2))
   ), use.names = TRUE
)
```



```{r write_germ_tm}
# save data into cache
fwrite(x = germ_tm[order(Tm,i_period, leaf),],
       file = "cache/germination_temperature.csv")

```

<br>  

***  

### Effect of temperature on branching  

We will do the same for data collected on the influence of temperature on the number of conidia which advanced past germination to branch and possibly develop infection structures.  

```{r tm_branch}
# read in data from sas file
br_tm <- readLines("data/tempvs.branching(results).sas")[3:40]
# cut out non-data lines
br_tm <- br_tm[-2]

#write it to csv so it can be read back in cand formatted to a data.frame instead of vector.
write(br_tm, "cache/Tm_branches.csv")

# read in data using fread
br_tm <- fread("cache/Tm_branches.csv", 
              col.names = c("obs", "Tm", "branches","rep", "conidia"))

# Add relative humidity and temperature data
br_tm[ , c("Tm","branches", "obs") := list(as.numeric(gsub("oC","",Tm)),
                                         as.numeric(gsub("b","",branches)),
                                         NULL)]

# recalculate vpd and convert branches to numeric
zero_branch <- br_tm[, (100 - sum(conidia)), 
                      by = .(Tm, rep)][,branches :=0][, .(Tm, branches, rep, V1)]

setnames(zero_branch,old = "V1", new = "conidia")

# Bind in the data containing non-germinated conidia as conidia with zero branches
br_tm <- rbind(br_tm,zero_branch)[, conidia := round(conidia)]
```

While not recorded in the raw data, no germination was observed at temperatures
8C or 35C. 
This is important to include for the data analysis.  

```{r}
br_tm <- rbindlist(
   list(
      br_tm,
      list(Tm = c(rep(8, 4 * 2),
                  rep(35, 4 * 2)),
           branches = rep(c(0,1,2,3), 
                         each = 2,
                         times = 2),
           rep = rep(1:2, 8),
           conidia = rep(c(100,0,0,0), 
                         each = 2,
                         times = 2)
      )
   ), use.names = TRUE
)
```

```{r save_br_tm}
fwrite(br_tm[order(Tm,branches, rep),], "cache/branching_temperature.csv")
```

<br>  

***  

## Experimental data on VPD  

### Effect of vpd on germination  


```{r clean_germ_vpd}
# Import
germ_vpd <- bind_rows(
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "B63:F158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  ),
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "I63:M158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  ),
  read_xls(
    "data/chapter 5 (germination vs VPD vs temp vs RH).xls",
    sheet = "Sheet1",
    range = "P63:T158",
    col_names = c("RH", "Tm", "i_period", "rep", "germ_n")
  )
)

# remove characters from temperature and time, make RH as a percentage
setDT(germ_vpd)[, c("RH", "Tm", "i_period","non_germ_n") :=
                  list(RH * 100,
                       as.numeric(gsub("C", "", Tm)),
                       as.numeric(gsub("h", "", i_period)),
                       100 - germ_n)]

# add vpd column
germ_vpd[, vpd := round(epiphytoolR::calc_vpd(RH = RH,
                                              Tm = Tm), 
                        3)]

# Change column order
setcolorder(germ_vpd,
            neworder = c("RH", "Tm", "vpd", "i_period", "rep", "germ_n", "non_germ_n"))
```




```{r save_g_vpd}
# write out clean data
fwrite(germ_vpd,
       file = "cache/germination_vpd.csv")
```

<br> 

***  

### Vapour pressure deficit effect on branching  
Read in the raw data.  

```{r}
# File not found 
br_vpd <-
  fread("data/220308_branching_vpd_temperature.csv",nrows = 29,header = TRUE)
           

```



```{r clean_br_vpd}
# Remove unnecessary columns and rows
br_vpd <- br_vpd[-c(5:10,15:24),-c(1,2,9:12,19:22,29)][-c(13)]

# rename leaf number column
setnames(br_vpd,
         old = 1:18,
         new = c("0.03_b1","0.13_b1","0.4_b1","0.7_b1","1.2_b1","1.8_b1",
                 "0.03_b2","0.13_b2","0.4_b2","0.7_b2","1.2_b2","1.8_b2",
                 "0.03_b3","0.13_b3","0.4_b3","0.7_b3","1.2_b3","1.8_b3"))

# Add temperature treatment
br_vpd[ , Tm := rep(c(22,25,28), each = 4)]

# reshape data to long format
br_vpd <-
   melt(br_vpd,
     id.vars = c("Tm"),
     value.name = "conidia",
     variable.factor = FALSE)

# split variable column into two columns vpd and branches and remove variable column
br_vpd[, c("vpd","branches") := tstrsplit(variable, "_b")][, variable := NULL]

# Change column order
setcolorder(br_vpd,
            neworder = c("vpd", "Tm", "branches","conidia"))

# set correct classes and add non-germinated
br_vpd[, c("vpd",
                "branches",
                "conidia",
                "rep") :=
              list(set_units(as.numeric(vpd),"kPa"),
                   as.numeric(branches),
                   as.numeric(conidia),
                   rep(1:4,.N/4))]

# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd[, (100 - sum(conidia)), 
                      by = .(Tm, vpd, rep)][,branches :=0][, .(vpd,Tm, branches, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Bind and roder data containing non-germinated conidia as conidia with zero branches
br_vpd <- rbind(br_vpd,zero_branch)[order(Tm,branches, rep)]


# save data into cache
fwrite(x = br_vpd,
       file = "cache/branching_VPD.csv")

```

### VPD branching dataset two  

```{r clean_br_vpd2}
br_vpd2 <- 
  read_xls("data/Chapter 5 (germination vs VPD vs temp vs RH).xls",
           sheet = "Sheet2",
           range = "AF3:AI182",
           col_names = c("vpd", "branches", "rep", "conidia"))

# make data.table format and add temperature and humidity
br_vpd2 <- data.table(br_vpd2)[, c("Tm","RH") := list(rep(c(22,25,28), each = 12, times = 5),
                                                            rep(c(95,85,75,55,32), each = 12*3))]

# recalculate vpd
br_vpd2[, vpd := round(epiphytoolR::calc_vpd(RH = RH,
                                                Tm = Tm),
                          3)]

# remove "b" from branches and make it numeric
br_vpd2[, branches := as.numeric(gsub("b","",branches))]

# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd2[, (100 - sum(conidia)), 
                      by = .(Tm, vpd, rep)][,branches :=0][, .(vpd,Tm, branches, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Change column order
setcolorder(br_vpd2[, RH := NULL],
            neworder = c("vpd", "Tm", "branches","rep","conidia"))

# Bind and roder data containing non-germinated conidia as conidia with zero branches
br_vpd2 <- rbind(br_vpd2,zero_branch)[order(Tm,branches, rep)]



# save data into cache
fwrite(x = br_vpd2,
       file = "cache/branching_VPD2.csv")

```


### More data on germtube branching under different vpd  

```{r clean_br_vpd3}
# read in data from sas file
dat <- readLines("data/vpd germtube.sas")[3:236]
# cut out non-data lines
dat <- dat[-c(2,55:58,111:114,167:170,223:226)]
#write it to csv
write(dat, "cache/vpd_branches.csv")

# read in data using fread
br_vpd3 <- fread("cache/vpd_branches.csv")

# Add relative humidity and temperature data
br_vpd3[ , c("RH","Tm") := list(rep(c(99,95,85,75,50,33),each = 36),
                                    rep(c(22,25,28),each = 12, times = 6))]

# recalculate vpd and convert branches to numeric
br_vpd3[, c("vpd",
                "bran") :=
              list(round(epiphytoolR::calc_vpd(RH = RH,
                                               Tm = Tm),
                         3),
                   as.numeric(gsub("b", "", bran)))]

# rearrange columns
setcolorder(br_vpd3,
            neworder = c("Obs", "RH", "Tm", "vpd", "bran", "rep", "numb")
            )

setnames(br_vpd3,
         old = c("bran", "numb"),
         new = c("branches","conidia"))


# Create data.table representing non-germinated conidia
## Assuming only 100 conidia were examined
zero_branch <- br_vpd3[, (100 - sum(conidia)), 
                      by = .(Tm, vpd, rep)][,branches :=0][, .(vpd,Tm, branches, rep, V1)]

# some treatment reps contained greater than 100 conidia assessments, therefore 
#  any negative number will be converted to zero and no non-germinated conidia are
#  assumed to be observed
zero_branch[,V1 := fifelse(V1 < 0, 0, V1)]

# rename column
setnames(zero_branch,old = "V1", new = "conidia")

# Change column order
setcolorder(br_vpd3[, c("RH","Obs") := .(NULL,NULL)],
            neworder = c("vpd", "Tm", "branches","rep","conidia"))

# Bind and roder data containing non-germinated conidia as conidia with zero branches
br_vpd3 <- rbind(br_vpd3, zero_branch)[order(Tm,branches, rep)]




fwrite(br_vpd3,
       file = "cache/branching_VPD3.csv")

```











