---
title: "The effect of temperature on *Podosphera xanthii* germination and branching"
description: |
  The statistical method undertaken in analysing the effect of temperature on 
  _P. xanthii_ germination and germ-tube production.  
author:
  - name: Paul Melloy
    url: https://PaulMelloy.github.io
    affiliation: The University of Queensland
    affiliation_url: https://uq.edu.au
date: "`r Sys.Date()`"
output: 
   distill::distill_article:
      toc: true 
      toc_depth: 3 
      toc_float: true 
---


## Compendium sections  

   1. [Title page](P_xanthii_titlepage.html)  
   2. [Data preparation and formatting](data_cleaning.html)  
   3. [The influence of temperatures on P. xanthii conidia germination](Pxanthii_temperature.html)  
   3. [The influence of VPD P. xanthii conidia germination and germtube formation](Pxanthii_vpd.html)  
   
   
<br>  
<br>  


Load R libraries required for analysis

```{r load_libraries, warning=FALSE, message=FALSE}
library(data.table) # data structure, cleaning and formatting
library(mgcv) # gam statistical models
library(rstanarm) # Bayesian statistical models
library(ggplot2) # graphing
library(MASS) # logistical ordinal regression functions, among other things
library(flextable) # formatting data.frames as word tables
library(marginaleffects) # produce predicted probabilities for polr()
source("R/p_star.R") # function to show visually the level of significance
```
<br>  

Read in data

```{r import_data}
g_tm <- fread("cache/germination_temperature.csv")
```



<br>  
<br>  

## Impact of Temperature on *P. xanthii* germination  
Peek at the data  

```{r data_peek}
head(g_tm)
str(g_tm)
```
<br>  

Summary table of the data.  

```{r}
g_tm_table <- 
   g_tm[, list(Mean = mean(germ_conidia),
               Std_dev = round(sd(germ_conidia),3),
               Median = median(germ_conidia),
            Range = paste0(range(germ_conidia), collapse = " - " ),
            Replicates = .N),
     by = .(Tm, i_period)] |>
   flextable() |>
   align(align = "center") |>
   set_header_labels(Tm = "Temperature",
                     i_period = "Incubation (hours)",
                     Std_dev = "Standard deviation") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   bg(i = seq(2,32,2),bg = "grey80",part = "body")|>
   set_caption("Podosheria xanthii germination percentage over a range of temperatures and incubation periods")
g_tm_table
```


```{r saveFig1, eval=FALSE, include=FALSE}
# Save table as a word document
doc <- officer::read_docx()
doc <- body_add_flextable(doc, value = g_tm_table)
#doc <- officer::body_end_section_landscape(doc)
print(doc, target = "FiguresTables/GermTemp_SummaryTable.docx")
```

How many leaves were inoculated?  

```{r}
length(unique(g_tm$Tm)) *           # Temperature incubators
   length(unique(g_tm$leaf)) *      # individual leaves
   length(unique(g_tm$i_period))    # Sampling times
```



### Visualise the raw data  
Let's plot the data over time, we will need to omit temperatures 8 and 35 where
no germination was observed.

```{r plot_raw_time}
g_tm[Tm != 8 &
        Tm != 35,] |>
   ggplot(aes(x = i_period, y = germ_conidia, colour = as.factor(Tm))) + 
   geom_point()+
   geom_smooth(method = "gam",formula = y ~ s(x, k = 3))+
   scale_color_viridis_d()+
   xlab("Time in hours")+
   ylab("Germinated conidia (%)")+
   labs(colour = "Temperature")+
   theme_classic()+
   scale_x_continuous(limits = c(12,48), breaks = c(12,18,24,30,36,42,48))+
   scale_y_continuous(limits = c(0,100), n.breaks = 5)
   
```
<br>  
Now lets plot germination over temperature separating by infection period.  

```{r plot_raw_Tm}
g_tm |>
   ggplot(aes(x = Tm, y = germ_conidia, colour = as.factor(i_period))) + 
   geom_point()+
   geom_smooth(method = "gam",formula = y ~ s(x, k = 8), alpha = 0.2)+
   scale_color_viridis_d(direction = -1)+
   xlab(expression("Temperature " ( degree*C)))+
   ylab("Germinated conidia (%)")+
   labs(colour = "Time (h)")+
   theme_classic()+
   scale_x_continuous(breaks = seq(5,35,5))+
   scale_y_continuous(limits = c(0,100), n.breaks = 5)+
   annotate("text", x = 8, y = 100, label = "(a)")

```


```{r saveFig2a, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig2a_TemperatureGermination.tiff",
   plot = Figure_2a,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```

This plot shows that the optimum temperature for rapid *P. xanthii* germination 
is around 28$^\circ$C. 
Also, in the first plot, if we assume at time zero there was no germinated conidia, 
50% of the conidia germinated within the first 12 hours. 
In all temperature treatments, between 12 and 36 hours after infection conidial 
germination increased at a linear rate. 
After 36 hours the rate of conidia germinating declined to almost zero.
Curves were fit using a generalised additive model with a smooth spline containing 6 knots.  

Lets further evaluate this fit.
However instead of fitting a temperature as a separate random effect, as depicted 
in the above plot, we will define temperature as a separate continuous predictor 
that may or may not interact with infection period.  

### Model analysis  

Germination is a binomial response so we need to provide the number of germinated 
to the number of non-germinated.
In the first model we will fit a smooth spline to each predictor and specify that each 
predictor influences germination independently of the other. 

I have used only 2 and 4 basis spline 'knots' for infection period and temperature
respectively.  

```{r model_1}
m1 <- gam(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k=3) + s(Tm, k =4), 
          data = g_tm,
          family = "binomial")
summary(m1)
plot(m1)
vis.gam(m1,
        theta = -50, # horizontal rotation
        phi = 30, # vertical rotation
        n.grid = max(g_tm$Tm) - min(g_tm$Tm) # grid cells
        )
```
<br>  
Model two we will investigate whether there is an interaction between temperature 
and infection period using a tensor.  

```{r model_2}
m2 <- gam(cbind(germ_conidia, non_germ_conidia) ~ te(i_period, Tm, k = 4),
          data = g_tm,
          family = "binomial")
summary(m2)
plot(m2)
vis.gam(m2, 
        theta = -50, # horizontal rotation
        phi = 30, # vertical rotation
        n.grid = max(g_tm$Tm) - min(g_tm$Tm) # grid cells
        )

```
This model shows slightly higher R squared and 'explained deviance' values, with 
a lower UBRE  value indicating that model two (`m2`) is the better model.  

Lets compare the models against each other.

```{r compare_m1m2}
anova(m1,m2, test = "Chisq")
```
This confirms that model two is the better model and explains more of the deviance.  

<br>  

Yet lets see if we can fit a better model to represent the raw data does not use 
an interaction to be sure this is the best fit. 
The plots of the fitted model indicate that 28$^\circ$C showed the highest conidial
germination. 
This might be due to the default type of basis splines, and number of knots. 
We have exhausted our options for improving the interaction, so let us investigate
improving the additive model by fitting a better spline to the temperature predictor.  

P-splines use a penalised least-square method to fit the model and are better in
non-normal distributions.
Lets try p-splines to see if we can improve the fit.

```{r model_3}
m3 <- gam(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k=3) + s(Tm, k =4, bs = "ps"),
          data = g_tm,
          family = "binomial")
summary(m3)
plot(m3)
vis.gam(m3,
        theta = -50, # horizontal rotation
        phi = 30, # vertical rotation
        n.grid = max(g_tm$Tm) - min(g_tm$Tm) # grid cells
        )
```
The P-splines have not improved the predictive ability of the model with a lower 
UBRE compared with model 2, improved the R squared value and explained deviance.
m3 however is a better model compared to m1 , 
We can also see that the curve has been shifted slightly up towards 28$^\circ$C.  


Lets compare the models two and three to against each other.

```{r comp_m2m3}
anova(m2,m3, test = "Chisq")
```
The models show no significant difference therefore we should use the simpler model,
model three (`m3`). 
Model three also shows a better fit to the data.  
  
<br>  
  
Next lets see if cyclic versions of the p-splines `"cp"` improve this fit further.

```{r model_4}
m4 <- gam(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k=3) + s(Tm, k =4, bs = "cp"),
          data = g_tm,
          family = "binomial")
summary(m4)
plot(m4)
vis.gam(m4,
        theta = -50, # horizontal rotation
        phi = 30, # vertical rotation
        n.grid = max(g_tm$Tm) - min(g_tm$Tm) # grid cells
        )
```
While not an improvement on `m2`, the cyclic P-splines seem to have a very good 
predictive ability with a very low UBRE, also a high R squared value and explaining 
a large amount of deviance.
We can also see that the curve has been shifted further towards 28$^\circ$C.  

<br>  
  
Lets compare models three and four.

```{r comp_m3m4}
anova(m2,m4, test = "Chisq")
anova(m3,m4, test = "Chisq")
```

<br>  

Lets use the p splines again however increase the number of knots for temperature
in attempt to get a better fit to the raw data.

```{r model_5}
m5 <- gam(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k=3) + s(Tm, k =6, bs = "cs"),
          data = g_tm,
          family = "binomial")
summary(m5)
plot(m5)
plot(m5,ylim = c(-3,4))
abline(v=28)
vis.gam(m5,
        theta = -50, # horizontal rotation
        phi = 30, # vertical rotation
        n.grid = max(g_tm$Tm) - min(g_tm$Tm) # grid cells
        )
```
This model plot looks contrived, however if we focus the plot and ignore large negative 
standardised values created by zero observations at 8 and 35$^\circ$, the curve 
seems to reflect the raw data very well. 
This model also shows the highest R-squared, explained more deviance and 
has the lowest UBRE score, compared to all other models.  


Lets compare the models two and five, then four and five to against each other.  

```{r comp_m4m5}
anova(m2,m5, test = "Chisq")
anova(m4,m5, test = "Chisq")
```
The cyclic p-spline model with more knots explained the most deviance and provides a 
significantly better fit.

And look at the AIC values of all the models.

```{r AIC_all}
AIC(m1,m2,m3,m4,m5)
```

This shows that model five is the best fit, with the lowest AIC and an R squared
of `r round(summary(m5)$r.sq, 4)`.
Temperature and infection period are significant predictors and the model produces
a low UBRE score or `r round(m5$gcv.ubre,3)`, indicating good predictive ability
and less affected by drop one, tests.

<br>  

### Visualise model against the raw data  

To simulate data from the model using `predict` we need to provide some values which
to estimate over.  

```{r}
# Make a data.frame of test values to feed in the model
test_dat <- data.table(Tm = rep(seq(from = 0, to = 40, by = 0.1), 4), # between 5 and 8 degrees for each incubation period
                        i_period = rep(c(12,24,36,48), each = length(seq(from = 0, to = 40, by = 0.1))))

# create a data.frame with predicted values and reference them to the respective temperature
m5_pred_list <-
   predict(m5,
           test_dat,
           type = "response",
           se.fit = TRUE)
m5_pred <- data.table(Tm = test_dat$Tm,
                      i_period = as.factor(test_dat$i_period),
                      germinated = m5_pred_list$fit,
                      germ_se = m5_pred_list$se.fit,
                      CI_low = m5_pred_list$fit + (m5_pred_list$se.fit * -1.94),
                      CI_high = m5_pred_list$fit + (m5_pred_list$se.fit * 1.94))

Figure_2a <- 
   m5_pred |>
   ggplot(aes(x = Tm, y = germinated, colour = i_period))+
   geom_line(size = 1)+
   scale_color_grey(start = 0.75,end = 0)+
   geom_point(data = g_tm, aes(x = Tm, y = (germ_conidia/100), colour = as.factor(i_period)))+
   theme_bw()+
   geom_line(aes(x = Tm, y = CI_low), linetype = "dashed", size = 0.6)+
   geom_line(aes(x = Tm, y = CI_high), linetype = "dashed", size = 0.6)+
   xlab(expression("Temperature " ( degree*C)))+
   ylab("Germinated conidia (%)")+
   labs(colour = "Time (h)")


Figure_2a
```

```{r saveFig2a, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig2a_TemperatureGermination.tiff",
   plot = Figure_2a,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```


***  

## Temperature effect on germ tube production  

Import cleaned data.  

```{r read_brTm_data}
br_tm <- fread("cache/branching_temperature.csv")
```


Peek at the data  

```{r br_data_peek}
head(br_tm)
str(br_tm)
summary(br_tm)
```

This experiment used the same incubation conditions as was used in the germination 
experiment. 
Following 36 hours of incubation at the respective temperatures, four leaf discs
were removed from the growth chamber stained and fixed for microscopy using the 
leaf clearing method. 
Then one hundred conidia were selected at random to record the state of germination 
or germ tube production.  

<br>  

### Visualise the raw data  

Summary table of the data.  

```{r}
br_tm[, list(Mean = mean(conidia),
            Range = paste0(range(conidia), collapse = " - " ),
            Std_dev = round(sd(conidia),3),
            Replicates = .N),
     by = .(Tm, germtubes)] |>
   flextable() |>
   align(align = "center") |>
   set_header_labels(Tm = "Temperature",
                     Std_dev = "Standard deviation") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   bg(i = seq(2,32,2),bg = "grey80",part = "body")|>
   set_caption("Podosheria xanthii germination percentage over a range of temperatures and incubation periods")

```


Let's plot the number of germ tubes for each temperature treatment. 
We can use stacked bars to show not only the number of germ tubes, but the fact that
a germinated conidia with three germtubes, also has two germ tubes ect.  

```{r plot_raw_br}
Figure_2b <-
   br_tm |>
   ggplot(aes( x = as.factor(Tm), y = conidia/2 , fill = factor(germtubes,levels = c(1,2,3,0)))) + 
   geom_col(position = position_stack(reverse = TRUE))+
   scale_fill_manual(values = c("0" = "grey90",
                                "1" = "grey70",
                                "2" = "grey55",
                                "3" = "grey40"), 
                     name = "Germ tubes")+
   xlab(expression("Temperature " ( degree*C)))+
   ylab("Germ tube production (%)")+
   labs(colour = "Germ tubes")+
   theme_classic()+
   annotate("text", x = 0.8, y = 99, label = "(a)", size = 5)

Figure_2b   
```

```{r saveFig2b, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig2b_TemperatureGermtube.tiff",
   plot = Figure_2b,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```

This plot shows that the optimum temperature for rapid *P. xanthii* germination 
is around 28$^\circ$C with `r br_tm[germtubes == 3 & Tm == 28, sum(conidia)]/2` % of 
conidia presenting three or more germ tubes. 
25$^\circ$C and 22$^\circ$C presented `r br_tm[germtubes == 3 & Tm == 25, sum(conidia)]/2` and 
`r br_tm[germtubes == 3 & Tm == 22, sum(conidia)]/2` % respectively.
Although there were only fewer conidia with tertiary germtubes in the 22°C treatment,
combined means of conidia with two or more germtubes were 
`r br_tm[germtubes >= 2 & Tm == 22, sum(conidia)]/2`,
`r br_tm[germtubes >= 2 & Tm == 25, sum(conidia)]/2`, and 
`r br_tm[germtubes >= 2 & Tm == 28, sum(conidia)]/2` in temperature treatments 
22$^\circ$C, 25$^\circ$C, and 28$^\circ$C.
Only `r br_tm[germtubes == 3 & Tm == 19, sum(conidia)]/2`% of germinated conidia 
had produced tertiary germtubes in the 19$^\circ$C treatment, with 
`r br_tm[germtubes >= 3 & Tm == 17, sum(conidia)]/2` tertiary germ tubes observed 
in either 17$^\circ$C or 31$^\circ$C treatments. 

Also, no germination was observed at temperatures 8$^\circ$C , or at 35$^\circ$C.

<br>  

### Ordinal logistic regression  

Lets build a model to help us predict germ tube production after 36 
hours when incubated at each respective temperature.  

Trecate (2019) claims if there are two or more germ tubes, successful infection is 
likely to have occurred.
Here we have a response variable, `germtubes` which describes the progress of a potential
infection by a single conidia.  
   - `0` which represents a non-germinated conidia,  
   - `1` a germinated conidia with one germ tube,  
   - `2` a germinated conidia with two germ tubes,  
   - `3` a germinated conidia with three or more germ tubes.  
   
The response variable might be thought of as continuous. 
However, given our aim is to estimate successful infection by the pathogen by proxy 
of germ tube production, estimating the number of germ tubes and hyphae 
beyond three is irrelevant and provide inaccurate results, given the difference between
the number of germtubes may not be equal or consistent between levels.
`germtubes` would be better characterised as ordinal and therefore leads us to using
*ordinal logistic regression* methods.  

The aim of this analysis is to predict the likelihood of a conidia germinating and 
branching under different temperature conditions.  
The data is count data and our response is not binomial but ordered multinomial, 
which leads us to use an ordinal logistic regression.  

First we need to convert the count data into a longer format, so each row represents
an observation.  

```{r to_long}
br_tm_lng <-
   rbindlist(apply(br_tm, 1 , function(x) {
      rbindlist(lapply(seq_len(x["conidia"]), function(x2) {
         as.data.frame(t(x[c(1:3)]))
      }))
   }))

# classify each column
br_tm_lng[, c("Tm", "germtubes", "rep") := list(Tm,
                                               ordered(germtubes), 
                                               as.factor(rep))]
```

#### Model 1 - temperature effect on germ-tube production  

Now we have prepared our data lets try it in our first model assuming a linear and 
additive relationship between germ tube number and temperature and leaf rep.  

```{r m1}
m1 <- polr(germtubes ~ Tm + rep, 
           data = br_tm_lng, 
           Hess = TRUE)

summary(m1)
```

The output is in proportional log odds and presents *Coefficients* which represent
the change in log odds for each unit of the predictor.
For example for each degree temperature increases the log odds increases `r round(coef(m1)[1],3)`.  

The coefficient `rep` is not significant and can probably be dropped from the model.  

Lets plot the model predictions for temperatures between 8$^\circ$ C and 35$^\circ$ C
to determine if the result is reasonable compared to the raw data.  

```{r plot_m1}
# Make a data.frame of test values to feed in the model
test_dat <- data.table(Tm = seq(from = 5, to = 40, by = 0.1), # between 5 and 8 degrees
                       rep = as.factor(rbinom(length(seq(from = 5, to = 40, by = 0.1)),1,0.5)+1)) # randomly use either the first or second rep

# create a data.frame with predicted values and reference them to the respective temperature
m1_pred <-
   as.data.table(cbind(Tm = test_dat$Tm,
                       round(predict(m1,
                                     test_dat,
                                     type = "p"),
                             3)))

# format to long 
m1_pred <-
   melt(
      m1_pred,
      id.vars = "Tm",
      measure.vars = c("0", "1", "2", "3"),
      variable.name = "germtubes",
      value.name = "prob"
   )

m1_pred |>
   ggplot(aes(Tm, prob, colour = germtubes))+
   geom_line(size = 1)+
   theme_minimal()
```

The result is rather linear and predicts more germtubes at 40C when the raw data does not see any from 35C or presumably above.

#### Model 2 - Temperature effect on germ-tube production  

Lets try applying some splines to temperature and dropping rep.  

```{r m2}
library(splines)
m2 <- polr(germtubes ~ bs(Tm,Boundary.knots = c(8,35)), 
                         data = br_tm_lng, Hess = TRUE)
summary(m2)
```
The residual deviance and AIC on this model is lower, indicating this might be a 
better fit. 
Also the t value is higher hinting this model shows clearer differences between 
the variables, especially in the middle of the studied temperature range.  

Now lets try predicting from this model.  

```{r pred_m2}
m2_pred <- as.data.table(round(predict(m2,
                         test_dat,
                         type = "p"),
                 3))[, Tm := test_dat$Tm]

```
We get a warning indicating the the predicted values are outside the range of the 
spline boundary knots. 
This is ok as we know there should be a zero probability of spore germination outside 
this temperature range of 8$^\circ$C and 35$^\circ$C

```{r}
# format to long 
m2_pred <-
   melt(
      m2_pred,
      id.vars = "Tm",
      measure.vars = c("0", "1", "2", "3"),
      variable.name = "germtubes",
      value.name = "prob"
   )

m2_pred |>
   ggplot(aes(Tm, prob, colour = germtubes))+
   geom_line(size = 1)+
   theme_minimal()
```

This plot looks like it would fit the raw data much better than `m1`.
However, the model estimates 25$^\circ$C as the temperature likely to produce
a greater proportion of germinated conidia with three germ-tubes. 
This does not reflect the raw data we plotted in figure 2b.  

#### Model 3 - Temperature effect on germ-tube production  

Lets see if we can improve the model by changing the number and placement of the 
spline knots.  

```{r m3}
m3 <- polr(germtubes ~ bs(Tm,
                         knots = c(19,23,27,30)), 
                         data = br_tm_lng, Hess = TRUE)
summary(m3)
```
The residual deviance and AIC on this model is lower, indicating this might be a 
better fit. 
Also the t value is higher hinting this model shows clearer differences between 
the temperature variables.  

Now lets try predicting from this model.  

```{r pred_m3}
m3_pred <- as.data.table(round(predict(m3,
                         test_dat[,1],
                         type = "p"),
                 3))[, Tm := test_dat$Tm]

```
We get a warning indicating the the predicted values are outside the range of the 
spline boundary knots. 
This is ok as we know there should be a zero probability of spore germination outside 
this temperature range of 8$^\circ$C and 35$^\circ$C

```{r plot_m3}
# format to long 
m3_pred <-
   melt(
      m3_pred,
      id.vars = "Tm",
      measure.vars = c("0", "1", "2", "3"),
      variable.name = "germtubes",
      value.name = "prob"
   )

m3_pred |>
   ggplot(aes(Tm, prob, colour = germtubes))+
   geom_line(size = 1)+
   theme_minimal()
```

The plot shows that this model might have too many spline knots. 
Lets compare the models using the `anova()` function.

```{r anova_m2-3}
anova(m2,m3)
```
While the anova says `m3` is a significantly better model, the residual deviance 
informs us it is not by much.  

#### Model 4 - Temperature effect on germ-tube production  

Based on the plot of model 3, we will try another model with fewer knots to reduce 
the wiggliness.  

```{r m4}
m4 <- polr(germtubes ~ bs(Tm,
                         knots = c(23,28)
                         ), 
                         data = br_tm_lng, Hess = TRUE)
summary(m4)
```
The residual deviance and AIC on this model is lower, indicating this might be a 
better fit. 
Also the t value is higher hinting this model shows clearer differences between 
the variables.  

Now lets try predicting from this model, we will use the .  

```{r pred_m4}
m4_pred <- marginaleffects::predictions(m4,
                                        newdata = test_dat[,1],
                                        type = "probs")
# set data.table format
setDT(m4_pred)

# calculate approximate confidence intervals
m4_pred[, c("CI_low","CI_high") := .(predicted - 1.96*std.error,
                                   predicted + 1.96*std.error)]

# change the column names
setnames(m4_pred,
         old = c("group","Tm.x"),
         new = c("germtubes", "Tm"))

```
We get a warning indicating the the predicted values are outside the range of the 
spline boundary knots. 
This is ok as we know there should be a zero probability of spore germination outside 
this temperature range of 8$^\circ$C and 35$^\circ$C

```{r plot_m4}
# format to long 
Figure_3b <- 
   m4_pred |>
   ggplot(aes(Tm, predicted *100, colour = germtubes))+
   geom_line(size = 1)+
   geom_line(data = m4_pred, aes(x = Tm, y = CI_low*100, #colour = group,
                                 ), linetype = "dotted", size = 0.7)+
   geom_line(data = m4_pred, aes(x = Tm, y = CI_high*100, #colour = germtubes,
                                 ), linetype = "dotted", size = 0.7)+
   theme_minimal()+
   scale_colour_viridis_d(direction = -1)+
   xlab(expression("Temperature " ( degree*C)))+
   ylab("Percentage (%)")+
   labs(colour = "Germ tubes")+
   geom_vline(aes(xintercept = 28), linetype = 2)+
   annotate(geom = "text",
            x = 29,
            y = 98,
            label = 28)+
   annotate(geom = "text",
            x = 6,
            y = 96,
            label = "(b)",size = 5)
Figure_3b
[```

```{r saveFig3b, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig3b_TemperatureGermtube.tiff",
   plot = Figure_3b,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```

The plot shows a smoother fit with the knots we have tried. 
Lets compare the model using the `anova()` function.

```{r anova_m3-4}
anova(m3,m4)
```
While the anova says `m3` is a significantly better model, the residual deviance 
informs us it is not by much. 
Because the `m3` plot looks over-fit and `m4` plot seems a more reasonable fit to the raw data, we will use this.  

Lets examine the coefficients.  

```{r m4_coef}
coef(m4)
```
There are five coefficients for the Temperature variable because the b-spline breaks
the data into different ranges depending on `knot` placement. 
Here we see there are four knots at the following temperatures

```{r}
tm_bs <- bs(br_tm_lng$Tm, knots = c(23,28))
sort(c(attr(tm_bs, which = "knots"),
  attr(tm_bs, which = "Boundary.knots")))
```
As can be seen in the summary above it is a little hard to make sense of these values. 
The spline breaks up the temperature predictor into five predictors, each explaining 
the difference between temperatures for a particular range.
The value gives us a scaled representation of the mean magnitude of each temperature
range on a log odds scale, and therefore the only inference we can make from these values is some are 
greater some are less and some have a large negative influence on spore germination
state.  

The significance of each predictor can be observed by the standard error or the 
't value' $t value = \frac{coef}{se}$. 
We can also obtain P-values from this output, however these methods
are not advised given they assume a normal distribution and infinite degrees of freedom.  

```{r m4_coefTab}
# get table of coefficients
coef_table <- data.table(coef(summary(m4)), keep.rownames = TRUE)

# calculate p-values
coef_table[, p_value := pnorm(abs(`t value`),lower.tail = FALSE) *2][, sig := p_star(p_value)]
coef_table
```

Perhaps a better method for inferring statistical significance is to consider the 
confidence intervals and if the CI includes zero.  

```{r m4_confit}
m4_ci <- confint(m4)
m4_ci
```

So for temperatures < 8$^\circ$C the coefficient was `r coef(m4)[1]`, with CIs that 
include zero indicating little difference between the number of conidial germ-tubes 
at temperature treatments in this range.  

For temperatures 8$^\circ$C < `Tm` < 23$^\circ$C the coefficient was `r coef(m4)[2]`, 
with CIs that _don't_ include zero indicating a likely significant difference between the
proportions of conidia with varying numbers of germ-tubes due to temperatures in 
this range.  

For temperatures 23$^\circ$C < `Tm` < 28$^\circ$C the coefficient was `r coef(m4)[3]`, 
with CIs that _don't_ include zero indicating a likely significant difference between the
proportions of conidia with varying numbers of germ-tubes due to temperatures in 
this range.  

For temperatures 28$^\circ$C < `Tm` < 35$^\circ$C the coefficient was `r coef(m4)[4]`, 
with CIs that _don't_ include zero indicating a likely significant difference between the
proportions of conidia with varying numbers of germ-tubes due to temperatures in 
this range.

For temperatures 35$^\circ$C < `Tm`  the coefficient was `r coef(m4)[5]`, 
with CIs that _don't_ include zero indicating a likely significant difference and negative 
association between the proportions of conidia with varying numbers of germ-tubes 
due to temperatures in this range.  



$\frac{t}{13.3}I(0<=t<13.3)+\frac{38.83333-t}{38.83333-13.3}I(13.3<=t<38.83333)$


