---
title: "The effect of vapour pressure deficit on *Podosphera xanthii* germination and germ tube production"
description: |
  The statistical method undertaken in analysing the effect of temperature on 
  _P. xanthii_ germination and germ-tube production.  
author:
  - name: Paul Melloy
    url: https://PaulMelloy.github.io
    affiliation: The University of Queensland
    affiliation_url: https://uq.edu.au
date: "`r Sys.Date()`"
output: 
   distill::distill_article:
      toc: true 
      toc_depth: 3 
      toc_float: true 
---
<style>
body {
text-align: justify}
</style>

## Compendium sections  

   1. [Title page](P_xanthii_titlepage.html)  
   2. [Data preparation and formatting](data_cleaning.html)  
   3. [The influence of temperatures on P. xanthii conidia germination](Pxanthii_temperature.html)  
   3. [The influence of VPD P. xanthii conidia germination and germtube formation](Pxanthii_vpd.html)  
   
***  

Load R libraries required for analysis

```{r load_libraries, warning=FALSE, message=FALSE}
library(data.table) # data structure, cleaning and formatting
library(mgcv) # gam statistical models
library(rstanarm) # Bayesian statistical models
library(ggplot2) # graphing
library(MASS) # logistical ordinal regression functions, among other things
library(splines) # non-linear predictors
library(flextable)
source("R/p_star.R")
```
<br>  

## Impact of vapour pressure deficit on *P. xanthii* germination  

Read in data

```{r import_data}
g_vpd <- fread("cache/germination_vpd.csv")
```

<br>  

Peek at the data

```{r data_peek}
str(g_vpd)
# inspect environmental treatments
unique(g_vpd[,.(RH,Tm,vpd)])
```
This table shows there are 18 VPD treatments consisting of six relative humidity
and three temperatures.  

Sampling times in hours.  

```{r}
g_vpd[, .(reps = paste0(range(rep), collapse = " - ")), by = i_period]
```

<br>  

### Visualise the raw data  
Let's plot the germination over time split by relative humidity.

```{r plot_raw_timexRH}
g_vpd |>
   ggplot(aes(x = i_period, y = germ_n, colour = as.factor(Tm))) + 
   geom_point()+
   facet_wrap(facets = "RH")+
   geom_smooth(method = "gam",formula = y ~ s(x, k = 3))+
   scale_color_viridis_d()+
   xlab("Time in hours")+
   ylab("Germinated conidia (%)")+
   labs(colour = "Temperature")+
   theme_classic() +
   ggtitle("Germinated conidia over time, subset by relative humidity")+
   scale_x_continuous(limits = c(12,48), breaks = c(12,24,36,48))+
   scale_y_continuous(limits = c(0,100), n.breaks = 5)
   
```
Germination increases in a linear fashion at all the relative humidity treatments
with exception to 99 % humidity, which shows a somewhat non-linear trend in the 22$^\circ$C
and 25$^\circ$C temperature treatments.  

Let's look at the same data in terms of VPD split by time. 

```{r plot_raw_timexVPD, fig.width=6, fig.height=5}
i_p_labs <-c("12 hours","24 hours","36 hours","48 hours")
names(i_p_labs) <- c(12,24,36,48)
g_vpd[, lty := fcase(Tm == 22, "dotted",
                         Tm == 25, "dashed",
                         Tm == 28, "twodash")]
Figure_4 <- 
  g_vpd |>
   ggplot(aes(x = vpd, y = germ_n,colour = factor(Tm))) + 
   geom_point()+
   geom_smooth(method = "gam",
               formula = y ~ bs(x,k=0.2),
               size = 0.7,se = FALSE)+
   scale_color_viridis_d()+
   facet_wrap(facets = c("i_period"), 
              labeller = labeller(i_period = i_p_labs))+
   xlab("Vapour pressure deficit (kPa)")+
   ylab("Germinated conidia (%)")+
   labs(colour = "Temperature")+
   theme_bw()+
   theme(strip.background = element_rect(fill = "white"))
Figure_4
```



```{r saveFig4, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig4_VPDGermination.tiff",
   plot = Figure_4,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```

Let's look at the same data in terms of VPD split by temperature. 

```{r plot_raw_TmxVPD,layout="l-body-outset",  fig.width=10, fig.height=5}
#i_p_labs <-c("12 hours","24 hours","36 hours","48 hours")
# names(i_p_labs) <- c(12,24,36,48)
g_vpd |>
   ggplot(aes(x = vpd, y = germ_n, colour = as.factor(i_period))) + 
   geom_point()+
   facet_wrap(facets = c("Tm"),nrow = 1)+
   geom_smooth(method = "gam",formula = y ~ bs(x,k=0.2), alpha = 0.3)+
   scale_color_viridis_d()+
   xlab("VPD")+
   ylab("Germinated conidia (%)")+
   labs(colour = "Incubation time")+
   theme_bw()+
   theme(strip.background = element_rect(fill = "white"))
```


This data is very hard to fit a curve to and seems like it would benefit with two functions.
Lets plot again without the saturated 99% humidity treatment.  

```{r plot_subset_rh}
i_p_labs <-c("12 hours","24 hours","36 hours","48 hours")
names(i_p_labs) <- c(12,24,36,48)
g_vpd[g_vpd$RH != 99] |>
   ggplot(aes(x = vpd, y = germ_n, colour = as.factor(Tm))) + 
   geom_point()+
   facet_wrap(facets = c("i_period"), 
              labeller = labeller(i_period = i_p_labs))+
   geom_smooth(method = "gam",formula = y ~ s(x,k=5), alpha = 0.1)+
   scale_color_viridis_d()+
   xlab("VPD")+
   ylab("Germinated conidia (%)")+
   labs(colour = "Temperature")+
   theme_classic()
```

From these plots it seems that the influence of VPD on germination is bimodal, with an increase in germination between 1.5 and 2 kPa. 
Particularly in the warmer treatments.  

***  


### VPD influence on germination - analysis

Lets start the analysis with a simple model with smoothing terms on VPD, temperature
and infection period.
We are using a binomial general additive model for all the germination analysis, 
because it is binary count data. 
As we saw from plotting the raw data, we likely have non-linear predictors and a 
`gam` model is a great way to estimate with splines on the predictors.  

```{r m1v}
m1v <- gam(cbind(germ_n,non_germ_n) ~ s(vpd, k = 5) + s(Tm, k = 3) + s(i_period, k = 3), 
           data = g_vpd,
           family = "binomial")
summary(m1v)
plot(m1v)
```

Overall the model fits very well with a R$^2$ of `r round(summary(m1v)$r.sq,4)`.
The splines on `Tm` and `i_period` were statistically significant, and seem to reflect
the raw data. 
Warmer temperatures, from the three temperatures included (22, 25 and 28$^\circ$C)
lead to more conidia germinating.
Incubation time also showed a positive relationship with the number of germinated 
conidia, ie. the longer the incubation the more conidia were likely to have germinated. 
VPD also was a significant predictor however the wiggliness might be able to be improved.  

As earlier, lets try and reduce the willingness by using P-splines.  

```{r m2v}
m2v <- gam(cbind(germ_n,non_germ_n) ~ s(vpd, k = 5, bs = "ps") + s(Tm, k = 3) + s(i_period, k = 3), 
           data = g_vpd,
           family = "binomial")
summary(m2v)
plot(m2v)
```

This shows only a small improvement and a decreasing trend in the VPD plot.  

Next lets try more VPD spline knots.  

```{r m3v}
m3v <- gam(cbind(germ_n,non_germ_n) ~ s(vpd, k = 7, bs = "ps") + s(Tm, k = 3) + s(i_period, k = 3), 
           data = g_vpd,
           family = "binomial")
summary(m3v)
plot(m3v)
```

While there exists a lot of willingness in the gam, the vpd plot shows the increase 
in germination around 1.5 kPa and a rapid decrease in germination between the saturated 
treatments (0.026 - 0.038 kPa), and the next highest humidity (0.158 - 0.189 kpa).
This model also provides a high R-squared (`r round(summary(m3v)$r.sq,4)`) and 
low UBRE `r round(m3v$gcv.ubre,3)`, indicating good predictive ability while being 
less affected by drop one, tests.

Temperature does not seem like it needs a non-linear term. The next model will 
test this assumption.  

```{r}
m4v <- gam(cbind(germ_n,non_germ_n) ~ s(vpd, k = 7) + Tm + s(i_period,k=3), 
           data = g_vpd,
           family = "binomial")
summary(m4v)
plot(m4v)
```

There is a slight reduction in the model fit, as shown by the difference in the 
R$^2$ and UBRE values between m3v and m4v.

In addition the wiggliness of the vpd curve seems more pronounced. 
We will next evaluate if there is an interaction between VPD and temperature, which
should also help us decide if temperature should be a linear or non-linear term.  


```{r}
m5v <- gam(cbind(germ_n,non_germ_n) ~ te(vpd,Tm, k = 3)+ s(i_period,k=3), 
           data = g_vpd,
           family = "binomial")
summary(m5v)
anova(m4v,m5v)
```
This comparison shows while the interaction between temperature is significant,
the m5v model does not fit the data as well as m4v .  

Because of the rapid drop in germination in the lower VPD range, and little change 
between the higher VPD treatments, it is difficult to fit a non-linear model with 
evenly spaced splines. 
Also we don't have enough evenly spaced VPD treatments to increase the number of 
knots without causing an over fit model. 
We can override this default behaviour of splines placed at set points by specifying 
our own knot points.  

Here we will specify two knots at 0.1 kPa and 0.4 kPa, in addition to the boundary
knots.  

```{r germ_bsmod}
b_k <- range(g_vpd$vpd) # set spline boundary knots
m6v <- gam(cbind(germ_n,non_germ_n) ~ s(vpd, k = 4,bs = "cs") + Tm + s(i_period,k=3),
           knots = list(vpd = c(b_k[1],0.1,0.4,b_k[2])),
           data = g_vpd,
           family = "binomial")


summary(m6v)
plot(m6v)

vis.gam(m6v,
        phi = 20,
        theta = 40,
        view = c("vpd","Tm"),
        )
```

This model presents the best fit model so far, as shown by the R$^2$ value of `r round(summary(m6v)$r.sq,4)`,
`r round(summary(m6v)$dev.exp,4)*100`% deviance explained and `r summary(m6v)$sp.criterion`UBRE.  

Let's compare the models side-by-side using the 'Akaike Information Criterion' (AIC).
A lower number is generally viewed as indicating a better model fit.  

```{r}
AIC(m1v,m2v,m3v, m4v, m5v, m6v)
```

Model six (`m6v`) shows the best fit to the raw data plotted in Figure 4.   

```{r}
# subset by incubation time
par(mfrow = c(2,2),
    oma = c(0,0,0,0),
    mar = c(0.1,0.1,1,0.1))
vis.gam(m6v,
        phi = 20,
        theta = 45,
        view = c("vpd","Tm"),
        cond = list(i_period = 12),
        main = "12 hours")
vis.gam(m6v,
        phi = 20,
        theta = 45,
        view = c("vpd","Tm"),
        cond = list(i_period = 24),
        main = "24 hours")
vis.gam(m6v,
        phi = 20,
        theta = 45,
        view = c("vpd","Tm"),
        cond = list(i_period = 36),
        main = "36 hours")
vis.gam(m6v,
        phi = 20,
        theta = 45,
        view = c("vpd","Tm"),
        cond = list(i_period = 48),
        main = "48 hours")

par(mfrow = c(1,1))
```

***  

## Impact of VPD on *P. xanthii* branching  

Import the clean data

```{r import_data_br}
# Read clean data
b_vpd <- fread("cache/branching_vpd.csv")[, exp := 1]
b_vpd2 <- fread("cache/branching_VPD2.csv")[,exp := 2]
b_vpd3 <- fread("cache/branching_VPD3.csv")[,exp := 3]
```

There were three data-sets which recorded germtube formation at a range of VPD. 

Lets look at the summaries of the data.  

```{r}
rmarkdown::paged_table(b_vpd[, list(
   Mean = mean(conidia),
   Std_dev = round(sd(conidia), 3),
   Median = median(conidia),
   Range = paste0(range(conidia), collapse = " - "),
   Replicates = .N),
   by = .(vpd, Tm, germtubes)])


rmarkdown::paged_table(b_vpd2[, list(
   Mean = mean(conidia),
   Std_dev = round(sd(conidia), 3),
   Median = median(conidia),
   Range = paste0(range(conidia), collapse = " - "),
   Replicates = .N),
   by = .(vpd, Tm, germtubes)])


rmarkdown::paged_table(b_vpd3[, list(
   Mean = mean(conidia),
   Std_dev = round(sd(conidia), 3),
   Median = median(conidia),
   Range = paste0(range(conidia), collapse = " - "),
   Replicates = .N),
   by = .(vpd, Tm, germtubes)])

```

<br>  

We will combine the data into one object to interrogate.  

```{r bv_combine}
b_v <-rbind(b_vpd,b_vpd2,b_vpd3)
```


### Visualise the raw data  
First we want observe the effect of VPD on the number of germtubes at 36 hours for
each experimental rep.  

```{r plot_br_vpd_exp, fig.dim=c(8,8)}
b_v[,percent := (conidia/sum(conidia))*100,
    by = .(vpd,exp)] 
b_v |>
   ggplot(aes( x = as.factor(vpd), 
               y = percent ,
               fill = as.factor(germtubes))) + 
   geom_col()+
   facet_grid(rows = vars(exp))+
   scale_fill_brewer(palette = "Blues", name = "Germ tubes")+
   xlab("Vapour pressure deficit")+
   ylab("Percent germinated")+
   labs(colour = "Germ tubes")+
   theme_classic()+
   theme(axis.text.x = element_text(angle = 70, vjust = 1.7, hjust = 2))+
   geom_text(aes(x = as.factor(vpd), y = 115, label = paste0(Tm,"°C")),size = 2.5)+
   scale_y_continuous(limits = c(0,120), breaks = c(0,25,50,75,100))
```
<br>  
Each experiment has similar trends and different variation for each vpd probably
due to the temperature incubation.  

Next lets look at a consolidated graph with all the experimental reps together.  

```{r plot_br_vpd}
b_v[,percent := (conidia/sum(conidia))*100,
    by = .(vpd)] 
   
Figure_5 <- 
   b_v |>
   ggplot(aes( x = as.factor(vpd), 
               y = percent ,
               fill = as.factor(germtubes))) + 
   geom_col()+
   scale_fill_brewer(palette = "Blues", name = "Germ tubes")+
   xlab("Vapour pressure deficit (kPa)")+
   ylab("Percent germinated (%)")+
   labs(colour = "Germ tubes")+
   theme_classic()+
   theme(axis.text.x = element_text(angle = 70, vjust = 1.7, hjust = 2))+
   geom_text(aes(x = as.factor(vpd), y = 105, label = paste0(Tm,"°C")),size = 2.5)+
   scale_y_continuous(limits = c(0,110), breaks = c(0,25,50,75,100))+
   annotate("text", x = 1, y = 110, label = "(a)", size =5)
Figure_5
```

```{r saveFig5, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig5_VPDGermtube.tiff",
   plot = Figure_5,
   device = "tiff",
   height = 6,
   width = 7,
   dpi = 300,
   units = "in"
)
```

What is the mean difference between the most saturated treatments 0.026 - 0.038 and
the next highest saturated treatments 0.132 - 0.186?  

```{r VPDiff}
# Temperature 22C
b_v[vpd == 0.026 &
       germtubes == 3, sum(percent)] -
   b_v[vpd == 0.132 &
       germtubes == 3, sum(percent)]

# Temperature 25C
b_v[vpd == 0.032 &
       germtubes == 3, sum(percent)] -
   b_v[vpd == 0.158 &
       germtubes == 3, sum(percent)]

# Temperature 28C
b_v[vpd == 0.038 &
       germtubes == 3, sum(percent)] -
   b_v[vpd == 0.189 &
       germtubes == 3, sum(percent)]

```


### Analysis of conidial germination and germtube proliferation
In order to analyse the data we need to convert the data to a longer format, where
each row represents an observation.

```{r model_1}
b_vpd_lng <-
   rbindlist(apply(b_v, 1 , function(x) {
      rbindlist(lapply(seq_len(x["conidia"]), function(x2) {
         as.data.frame(t(x[c(1:4,6,7)]))
      })
      )
   }))

b_vpd_lng[, c("exp",
              "rep",
              "germtubes",
              "cham",
              "vpd") := list(as.integer(exp),
                             as.integer(rep),
                             as.ordered(germtubes),
                             paste0(Tm, RH),
                             as.numeric(vpd))]

```

Test change

```{r}
m1 <- polr(germtubes ~ vpd + Tm+exp, data = b_vpd_lng, Hess = TRUE)

summary(m1)
```

Lets try predicting from this model so we can plot how it performs.
First we need to make some test data.

```{r test_dat}
# Make a data.frame of test values to feed in the model
test_dat <- data.table(Tm = rep(seq(from = 22, to = 28, by = 1),
                                each = 256,
                                times = 3), # between 5 and 8 degrees
                       vpd = rep(seq(from = 0.020, to = 2.57, by = 0.01),
                                 times = 7*3),
                       exp = rep(1:3,
                                 each = 7*256)
                       )
```

```{r, fig.dim=c(7,10)}
# create a data.frame with predicted values and reference them to the respective temperature
m1_pred <-
   as.data.table(cbind(test_dat,
                       round(predict(m1,
                                     test_dat,
                                     type = "p"),
                             3)))

# format to long 
m1_pred <-
   melt(
      m1_pred,
      id.vars = c("Tm","vpd","exp"),
      measure.vars = c("0", "1", "2", "3"),
      variable.name = "germtubes",
      value.name = "prob"
   )

m1_pred |>
   ggplot(aes(x = vpd, y = prob, colour = germtubes))+
   facet_grid(rows = vars(Tm), cols = vars(exp))+
   geom_line(size = 1)+
   theme_minimal()
```
There is a lot going on in this graph but as the `m1` summary shows experimental rep,
temperature and VPD were all significant. For ease in interpretation, lets look at 
a consolidated plot.  

```{r}
m1_pred[,.(prob = mean(prob)),by= .(vpd,germtubes)] |>
   ggplot(aes(x = vpd, y = prob, colour = germtubes))+
#   facet_wrap(facets = vars(Tm))+
   geom_line(size = 1)+
   theme_minimal()
```


<br>  
Model two we will investigate whether there is an interaction between temperature 
and infection period.  

```{r model_2}
m2 <- polr(germtubes ~ bs(vpd,knots = 0.5) + Tm + exp, data = b_vpd_lng, Hess = TRUE)
summary(m2)
```
Let's predict from this model to consider it's performance.


```{r}
# create a data.frame with predicted values and reference them to the respective temperature
m2_pred <-
   as.data.table(cbind(test_dat,
                       round(predict(m2,
                                     test_dat,
                                     type = "p"),
                             3)))

# format to long 
m2_pred <-
   melt(
      m2_pred,
      id.vars = c("Tm","vpd","exp"),
      measure.vars = c("0", "1", "2", "3"),
      variable.name = "germtubes",
      value.name = "prob"
   )

m2_pred[,.(prob = mean(prob)),by= .(vpd,Tm,germtubes)] |>
   ggplot(aes(x = vpd, y = prob, colour = germtubes))+
   facet_wrap(facets = vars(Tm))+
   geom_line(size = 1)+
   theme_minimal()
```

There is a consistent effect of temperature that increases the number of germinated
and branching germ-tubes. 
Low VPD has a strong effect on the number and proportion of branching germtubes.
While the number of germinated conidia and germtube branching decreases rapidly as 
VPD increases, however after 0.5 kPa, the proportion of germinating and branching 
conidia stabilise all the way up to 2 kPa.  

Lets look at just VPD and drop the temperature effect.

```{r}
Figure_5b <- 
   m2_pred[,.(prob = mean(prob)),by= .(vpd,germtubes)] |>
   ggplot(aes(x = vpd, y = prob, colour = germtubes))+
   #facet_wrap(facets = vars(Tm))+
   geom_line(size = 1)+
   theme_minimal() + 
   annotate("text", x = 0.01, y = 0.98, label = "(b)", size =5)+
   scale_colour_viridis_d(direction = -1)+
   xlab("Vapour pressure deficit (kPa)")+
   ylab("Proportional probability")+
   labs(colour = "Germ tubes")
Figure_5b
```

```{r saveFig5b, eval=FALSE, include=FALSE}
# Save table as a word document
ggsave(
   filename = "FiguresTables/Fig5b_VPDGermtubeMod.tiff",
   plot = Figure_5b,
   device = "tiff",
   height = 5,
   width = 7,
   dpi = 300,
   units = "in"
)
```


Given the plots show a decent similarity to the raw data we will choose `m2`.
Lets now extract the coefficients

```{r m2_coefTab}
# get table of coefficients
coef_table <- data.table(coef(summary(m2)), keep.rownames = TRUE)

# calculate p-values
coef_table[, p_value := pnorm(abs(`t value`),lower.tail = FALSE) *2][, sig := p_star(p_value)]
coef_table
```



## Baysian approach  

```{r install_stan, include=FALSE}
#install.packages("rstanarm")
library(rstanarm)
library(bayesplot)
```

```{r, include=FALSE}
# fit1 <- stan_gamm4(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k = 3) + s(Tm, k = 6, bs = "ps"), 
#                    data = g_tm, family = "binomial", chains = 1)
# 
# plot_nonlinear(fit1)
# 
# m5 <- gam(cbind(germ_conidia, non_germ_conidia) ~ s(i_period, k=3) + s(Tm, k =4, bs = "ps"),
#           data = g_tm,
#           family = "binomial",
# )
# 
# prior_summary(fit1)
# summary(fit1)
# mcmc_areas(as.matrix(fit1),
#            pars = "Tm",
#            prob = 0.95)
# pp_check(fit1,"stat")
```


```{r, include=FALSE}
# g_vpd %>%
#   mutate(germ = germ_n/(non_germ_n + germ_n)) %>%
#   group_by(vpd,i_period) %>%
#   summarise(m_germ = mean(germ),
#             Tm= unique(Tm)) %>%
#    filter(i_period == 12 &
#              vpd > 0.13) %>%
#    summary()
```
